---
title: "PCA_ANT3814"
output: html_document
date: "2024-06-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> Load library

```{r}
library(ggplot2)

```

> Cleaning up the data

# Unfortunately, we need to do a bit of work to get our data into reasonable shape. First we will remove a nuisance column (plink outputs the individual ID twice). We will also give our pca data.frame proper column names.

```{r}

# load the eigenvalues and eigenvectors from PLINK

pca <- read.table("~/PCA/cichlids.eigenvec")
eigenval <- scan("~/PCA/cichlids.eigenval")

# Remove the duplicated column 
pca <- pca[,-1]

# give the columns new names. 
names(pca)[1] <- "Individual" 
names(pca)[2:ncol(pca)] <- paste0("PC", 1:(ncol(pca)-1))

```

> Next we can add new columns for the species and sampling locations. This information is already contained in the "Individual" column, but it is messy. We can separate the information into separate columns using the R version of grep. We then use paste0 to combine the columns.

```{r}

# separate the individual species and populations

# Species

species <- rep(NA, length(pca$Individual))
species[ grep("PunPund", pca$Individual)] <- "pundamilia"
species[ grep("PunNyer", pca$Individual)] <- "nyererei"

# Location

location <- rep(NA, length(pca$Individual))
location[grep("Mak", pca$Individual)]  <- "makobe"
location[grep("Pyt", pca$Individual)]  <- "python"

```

> With these variables created, we can remake our data.frame by adding the new columns onto the end of the pca data frame.like so.

```{r}

# remake data frame

pca <- data.frame(pca, species, location)

```

> Plotting the data

# Now that we have done our data cleaning, we have everything in place to visualise the data properly. First we will plot the eigenvalues to see the relative importance of each. It is relatively straightforward to translate these into percentage variance explained.

```{r}

# Convert the eigenvectors to the percentage of the variance explained

pve <- data.frame(PC = 1:16, pve = eigenval/sum(eigenval)*100)

```

# With that done, it is simple to create a bar plot showing the percentage of the total variance that each principal component explains


```{r}

# make bar plot

ggplot(pve, aes(x = PC, y = pve)) +
  geom_bar(stat = 'identity') +
  ylab("Percentage of variance explained")+
  theme_light()

```

# Cumulatively, all 16 eigenvalues explain 100% of the variance, but PC1, PC2, and possibly PC3 together explain about 30% of the variance

> Plot the PCA

# Next we move on to actually plotting our PCA. Given the work we did earlier to reformat the data, this doesn't take much effort. 

```{r}

# plot PCA

ggplot(pca, aes(x= PC1, y = PC2, color = species, shape = location)) +geom_point(size =3)+
  xlab(paste0("PC1 (", signif(pve$pve[1], 3), "%)")) +
  ylab(paste0("PC2 (", signif(pve$pve[2], 3), "%)")) +
  theme_light()

```

> QUESTIONS 

#1: How much of the total variance is explained by the first and second principal components?

#2: Which variable is separated best by PC1 and which variable is best separated by PC2? 

#3: Does species explain more of the variance in the data or does sampling location?

#4: Which group of samples clusters separately from the other three groups?

#5: Remake the plot, but this time plut PC 2 on the X-axis and PC 3 on the Y axis. Paste the ggplot code here

#6: When removing the first principal component, are you able to more cleary see the separtion by species? Does this change your interpretation of the data? 

#7: There are 16 total samples in the data set and 4 groups. Is this enough data to clearly identify the separation of 2 variables? If you were designing a new study of these fish, what would you recommend doing to get a better understanding of their population genetic structure?   